---
title: "MantelProcrustes"
author: "Arianna Krinos"
date: "9/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/akrinos/Documents/Virginia Tech/Belden Lab/newt-long-term-2020")
setwd("/Users/akrinos/Documents/Virginia Tech/Belden Lab/newt-long-term-2020")
```

## Walking through Mantel/Procrustes

```{r downloads, include=FALSE}
pacman::p_load(devtools,tidyverse,stringr,qdapRegex,reshape2,plyr,
               RColorBrewer,vegan,proxy,philentropy,ade4,ape)
```

Start by reading in microbiome and metabolite data:

```{r read}
metabolite = read.csv("./data/metabolite.csv")
tomelt = read.csv("./data/LTEE_Newt_Seasonal_OTU_table_97_forprimer.csv")
tomelt = subset(tomelt, select = c(X.OTU.ID, AmphibID, Timepoint, Date, 
                                   Substrate.Addition))

# we only consider the timepoints/newts with both metabolite & microbiome data
metabolite = metabolite %>% left_join(tomelt, by = c("NewtID"="AmphibID",
                                                     "Timepoint", "Date")) %>%
  tidyr::drop_na(X.OTU.ID)
otulist = unique(as.character(metabolite$X.OTU.ID))
disturbance = as.character(metabolite$Substrate.Addition)

metabolite = subset(metabolite, select = -c(NewtID,	Timepoint,	Date,
                                            Month,	Season,	Year,	Season_Year,
                                            Month_Year,	Temp,	DO,	pH,	count,
                                            X.OTU.ID,Substrate.Addition))

microbiome = read.csv("./data/LTEE_Newt_Seasonal_2.csv")
microbiome = microbiome[,as.character(colnames(microbiome)) %in% otulist]
microbiome = microbiome[,otulist]
microbiome = t(microbiome)
```

Now we use `vegan` to calculate distance matrices:

```{r distances}
metabolitedist = vegdist(metabolite, method="jaccard",na.rm=TRUE)
microbiomedist = vegdist(microbiome, method="jaccard",na.rm=TRUE)
```

Let's look at the matrices and make sure they look alright: 

```{r examine}
#print(head(as.matrix(metabolitedist)))
#print(head(as.matrix(microbiomedist)))
```


That's what we expect. Now let's perform a Mantel test. 

But first, let's put this whole process into a function, so that we can do it again easily, and also subset if we want:

```{r createfunction}
processMetMic <- function(dist_metric="jaccard",substratefilter="") {
  metabolite = read.csv("./data/metabolite.csv")
  tomelt = read.csv("./data/LTEE_Newt_Seasonal_OTU_table_97_forprimer.csv")
  tomelt = subset(tomelt, select = c(X.OTU.ID, AmphibID, Timepoint, Date, 
                                     Substrate.Addition))
  
  # we only consider the timepoints/newts with both metabolite & microbiome data
  if (substratefilter == "") {
    metabolite = metabolite %>% left_join(tomelt, by = c("NewtID"="AmphibID",
                                                         "Timepoint", "Date")) %>%
      tidyr::drop_na(X.OTU.ID)
  } else {
    metabolite = metabolite %>% left_join(tomelt, by = c("NewtID"="AmphibID",
                                                       "Timepoint", "Date")) %>%
    tidyr::drop_na(X.OTU.ID) %>%
    dplyr::filter(as.character(Substrate.Addition) == substratefilter) 
    # choose from "Pre substrate addition", "Post substrate addition 1", 
    # and "Post substrate addition 2"
  }
  otulist = unique(as.character(metabolite$X.OTU.ID))
  
  metabolite = subset(metabolite, select = -c(NewtID,	Timepoint,	Date,
                                              Month,	Season,	Year,	Season_Year,
                                              Month_Year,	Temp,	DO,	pH,	count,
                                              X.OTU.ID,Substrate.Addition))
  rownames(metabolite) = otulist
  
  microbiome = read.csv("./data/LTEE_Newt_Seasonal_2.csv")
  microbiome = microbiome[,as.character(colnames(microbiome)) %in% otulist]
  microbiome = microbiome[,otulist]
  microbiome = t(microbiome)
  
  metabolitedist = vegdist(metabolite, method=dist_metric,na.rm=TRUE)
  microbiomedist = vegdist(microbiome, method=dist_metric,na.rm=TRUE)
  return(list("microbiomedist" = microbiomedist, "metabolitedist" = metabolitedist))
}
```

Let's start with a Mantel test on the full dataset: 

```{r mantelfull}
results = processMetMic(dist_metric="jaccard",substratefilter = "")
microbiomedist = results$microbiomedist
metabolitedist = results$metabolitedist
mantel.rtest(metabolitedist, microbiomedist, nrepet = 999)
#mantel.test(metabolitedist, microbiomedist, nperm = 999, alternative = "two.sided")

mantel(metabolitedist, microbiomedist, method="pearson", permutations=999, 
       strata = NULL,
    na.rm = FALSE, parallel = getOption("mc.cores"))
```

We can see that on the full dataset, the Mantel test is marginally significant (some stochastic samples end up with p < 0.05, some are slightly over 0.05). Let's try a Procrustes rotation and see whether it is significant: 

```{r procrustes}
# Do a MDS on the distance matrices
microbiomeMDS <- prcomp(microbiomedist) #monoMDS(microbiomedist) model = c("global","local","linear","hybrid")
metaboliteMDS <- prcomp(metabolitedist) #monoMDS(metabolitedist)
# Perform Procrustes
vare.proc <- procrustes(microbiomeMDS, metaboliteMDS, symmetric=FALSE)
summary(vare.proc)
plot(vare.proc)
plot(vare.proc, kind=2)
residuals(vare.proc)
print(protest(microbiomeMDS,metaboliteMDS, scores = "sites", symmetric=FALSE,
        permutations = how(nperm = 999)))
protest_res = protest(microbiomeMDS,metaboliteMDS, scores = "sites", symmetric=FALSE,
        permutations = how(nperm = 999))

ctest <- data.frame(rda1=vare.proc$Yrot[,1],
                    rda2=vare.proc$Yrot[,2],
                    xrda1=vare.proc$X[,1],
                    xrda2=vare.proc$X[,2],
                    dates=disturbance)

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
all = ggplot(ctest) +
  geom_point(aes(x=rda1, y=rda2, colour=dates)) +
  geom_point(aes(x=xrda1, y=xrda2, colour=dates)) +
  geom_segment(aes(x=rda1,y=rda2,xend=xrda1,yend=xrda2,colour=dates),
               arrow=arrow(length=unit(0.2,"cm"))) +
  theme_classic() + 
  xlab("Principal Component 1") + 
  ylab("Principal Component 2") + 
  scale_color_manual(name = "Disturbance",values=colorBlindBlack8[1:3],
                                breaks = c("Pre substrate addition",
                                "Post substrate addition 1",
                                "Post substrate addition 2"),
                     labels=c("PreD","PostD1","PostD2")) + 
  ggtitle(paste0("Full matrix, p=",as.character(round(protest_res$signif,4))))
all

ggsave("procrustesrotation_all_9Oct.jpg")
```

So we can see that the test of significance on the Procrustes rotation is not significant. Let's do the same thing, but write a function to print everything out first:

```{r writeafunct}
procrustesfunct <- function(metabolitedist, microbiomedist) {
  print(mantel(metabolitedist, microbiomedist, permutations = 999))
  # Do a MDS on the distance matrices
  microbiomeMDS <- prcomp(microbiomedist) #monoMDS(microbiomedist)
  metaboliteMDS <- prcomp(metabolitedist) #monoMDS(metabolitedist)
  # Perform Procrustes
  vare.proc <- procrustes(microbiomeMDS, metaboliteMDS)
  summary(vare.proc)
  plot(vare.proc)
  plot(vare.proc, kind=2)
  residuals(vare.proc)
  print(protest(microbiomeMDS,metaboliteMDS, scores = "sites", 
          permutations = how(nperm = 999)))
  protest_res = protest(microbiomeMDS,metaboliteMDS, scores = "sites", 
          permutations = how(nperm = 999))
  return(list("vp" = vare.proc,"signif" = protest_res$signif))
}
```

Now let's subset to pre-substrate addition:

```{r subset1}
results = processMetMic(substratefilter = "Pre substrate addition")
microbiomedist = results$microbiomedist
metabolitedist = results$metabolitedist
pcres = procrustesfunct(metabolitedist, microbiomedist)
vare.proc = pcres$vp
signif = pcres$signif

ctest <- data.frame(rda1=vare.proc$Yrot[,1],
                    rda2=vare.proc$Yrot[,2],
                    xrda1=vare.proc$X[,1],
                    xrda2=vare.proc$X[,2],
                    dates="Pre substrate addition")

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
pre = ggplot(ctest) +
  geom_point(aes(x=rda1, y=rda2, colour=dates)) +
  geom_point(aes(x=xrda1, y=xrda2, colour=dates)) +
  geom_segment(aes(x=rda1,y=rda2,xend=xrda1,yend=xrda2,colour=dates),
               arrow=arrow(length=unit(0.2,"cm"))) +
  theme_classic() + 
  xlab("Principal Component 1") + 
  ylab("Principal Component 2") + 
  scale_color_manual(name = "Disturbance",values=colorBlindBlack8[1],
                                breaks = c("Pre substrate addition"),
                     labels=c("PreD"))+ 
  ggtitle(paste0("Pre-disturbance, p=",as.character(round(signif,4))))
pre 
ggsave("procrustesrotation_pre.jpg")
```

So none of the tests are significant with only pre-substrate addition data. How about post disturbance 1? 

```{r subset2}
results = processMetMic(substratefilter = "Post substrate addition 1")
microbiomedist = results$microbiomedist
metabolitedist = results$metabolitedist
pcres = procrustesfunct(metabolitedist, microbiomedist)
vare.proc = pcres$vp
signif = pcres$signif


ctest <- data.frame(rda1=vare.proc$Yrot[,1],
                    rda2=vare.proc$Yrot[,2],
                    xrda1=vare.proc$X[,1],
                    xrda2=vare.proc$X[,2],
                    dates="Post substrate addition 1")

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
post1 = ggplot(ctest) +
  geom_point(aes(x=rda1, y=rda2, colour=dates)) +
  geom_point(aes(x=xrda1, y=xrda2, colour=dates)) +
  geom_segment(aes(x=rda1,y=rda2,xend=xrda1,yend=xrda2,colour=dates),
               arrow=arrow(length=unit(0.2,"cm"))) +
  theme_classic() + 
  xlab("Principal Component 1") + 
  ylab("Principal Component 2") + 
  scale_color_manual(name = "Disturbance",values=colorBlindBlack8[2],
                                breaks = c("Post substrate addition 1"),
                     labels=c("PostD1"))+ 
  ggtitle(paste0("Post disturbance 1, p=",as.character(round(signif,4))))
post1
ggsave("procrustesrotation_post1.jpg")
```

Or post disturbance 2?

```{r subset3}
results = processMetMic(substratefilter = "Post substrate addition 2")
microbiomedist = results$microbiomedist
metabolitedist = results$metabolitedist
pcres = procrustesfunct(metabolitedist, microbiomedist)
vare.proc = pcres$vp
signif = pcres$signif

ctest <- data.frame(rda1=vare.proc$Yrot[,1],
                    rda2=vare.proc$Yrot[,2],
                    xrda1=vare.proc$X[,1],
                    xrda2=vare.proc$X[,2],
                    dates="Post substrate addition 2")

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
post2 = ggplot(ctest) +
  geom_point(aes(x=rda1, y=rda2, colour=dates)) +
  geom_point(aes(x=xrda1, y=xrda2, colour=dates)) +
  geom_segment(aes(x=rda1,y=rda2,xend=xrda1,yend=xrda2,colour=dates),
               arrow=arrow(length=unit(0.2,"cm"))) +
  theme_classic() + 
  xlab("Principal Component 1") + 
  ylab("Principal Component 2") + 
  scale_color_manual(name = "Disturbance",values=colorBlindBlack8[3],
                                breaks = c("Post substrate addition 2"),
                     labels=c("PostD2"))+ 
  ggtitle(paste0("Post disturbance 2, p=",as.character(round(signif,4))))
post2

ggsave("procrustesrotation_post2.jpg")
```

Or post either disturbance? 

```{r subset4}
metabolite = read.csv("./data/metabolite.csv")
tomelt = read.csv("./data/LTEE_Newt_Seasonal_OTU_table_97_forprimer.csv")
tomelt = subset(tomelt, select = c(X.OTU.ID, AmphibID, Timepoint, Date, 
                                   Substrate.Addition))

# we only consider the timepoints/newts with both metabolite & microbiome data
metabolite = metabolite %>% left_join(tomelt, by = c("NewtID"="AmphibID",
                                                   "Timepoint", "Date")) %>%
tidyr::drop_na(X.OTU.ID) %>%
dplyr::filter((as.character(Substrate.Addition) == "Post substrate addition 1") | 
              (as.character(Substrate.Addition) == "Post substrate addition 2") )
otulist = unique(as.character(metabolite$X.OTU.ID))
substrates = metabolite$Substrate.Addition

metabolite = subset(metabolite, select = -c(NewtID,	Timepoint,	Date,
                                            Month,	Season,	Year,	Season_Year,
                                            Month_Year,	Temp,	DO,	pH,	count,
                                            X.OTU.ID,Substrate.Addition))

microbiome = read.csv("./data/LTEE_Newt_Seasonal_2.csv")
microbiome = microbiome[,as.character(colnames(microbiome)) %in% otulist]
microbiome = microbiome[,otulist]
microbiome = t(microbiome)

metabolitedist = vegdist(metabolite, method="jaccard",na.rm=TRUE)
microbiomedist = vegdist(microbiome, method="jaccard",na.rm=TRUE)
pcres = procrustesfunct(metabolitedist, microbiomedist)
vare.proc = pcres$vp
signif = pcres$signif


ctest <- data.frame(rda1=vare.proc$Yrot[,1],
                    rda2=vare.proc$Yrot[,2],
                    xrda1=vare.proc$X[,1],
                    xrda2=vare.proc$X[,2],
                    dates=substrates)

colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
postboth = ggplot(ctest) +
  geom_point(aes(x=rda1, y=rda2, colour=dates)) +
  geom_point(aes(x=xrda1, y=xrda2, colour=dates)) +
  geom_segment(aes(x=rda1,y=rda2,xend=xrda1,yend=xrda2,colour=dates),
               arrow=arrow(length=unit(0.2,"cm"))) +
  theme_classic() + 
  xlab("Principal Component 1") + 
  ylab("Principal Component 2") + 
  scale_color_manual(name = "Disturbance",values=colorBlindBlack8[2:3],
                                breaks = c("Post substrate addition 1",
                                "Post substrate addition 2"),
                     labels=c("PostD1","PostD2"))+ 
  ggtitle(paste0("Post disturbance, p=",as.character(round(signif,4))))
postboth

ggsave("procrustesrotation_postboth.jpg")
```

```{r procrustesplots}
cowplot::plot_grid(all+ggtitle("Full matrix"),
                   pre+ggtitle("Pre-disturbance"),
                   post1+ggtitle("Post-disturbance 1"),
                   post2+ggtitle("Post-disturbance 2"),
                   postboth+ggtitle("Post-disturbance"),ncol=3,labels = c("A","B","C","D","E"))
ggsave("mantelfigure.pdf",width=11,height=5,units="in",dpi=800)
ggsave(file="figure7.tiff",dpi=500,width=15,height=10,units="in")
```